# Chapter 5: Dimensionality reduction

load required libraries
```{r}
library(tidyverse)
library(psych)
library(corrplot)
library(tibble)
library(FactoMineR)
```


### Part 1: Overview

Read the data on transpose columns to rows

```{r}
human = read.csv('data/human.csv',header = TRUE)
human_ <- column_to_rownames(human, "Country")
```

Check the distributions and correlations between variables

```{r}
describe(human)
cor(human_)
corrplot(cor(human_))
hist(human_$GNI)
hist(human_$Mat.Mor)
```




GNI and maternal mortality are substantially skewed right. The rest of the variables are reasonably normal regarding skewness and kurtosis values. 

Maternal mortality and Adolescent birthrate correlate positively with each other and negatively with Life expectancy, Education, GNI and ratio of secondary education between genders, which form another intercorrelated cluster.

Representation in Parliament and Labor role participation ratio between genders do not correlate with other variables.


### Part 2: PCA on raw data

Calculate and visualize PCA

```{r}
pca_human <- prcomp(human_)
s <- summary(pca_human)
round(1*s$importance[2, ], digits = 2)
biplot(pca_human, cex = c(0.5, 0.8), col = c("grey40", "deeppink2"))
```

### Part 3: PCA on scaled data

Calculate and visualize PCA using normalized variables

```{r}
pca_human_z <- prcomp(scale(human_))
s_z <- summary(pca_human_z)
pca_pr_z <- round(100*s_z$importance[2, ], digits = 2)
pca_pr_z
pc_lab<-paste0(names(pca_pr_z), " (", pca_pr_z, "%)")
biplot(pca_human_z, cex = c(0.4, 0.6), col = c("grey40", "deeppink2"),xlab=pc_lab[1], 
       ylab=pc_lab[2])
```

*Caption for Figure: PC1 can be interpreted to represent the (negative) development level and prosperity of the society, PC2 represents gender equality. Numbers in parentheses represent accounted variance by the component*


With unstandardized variables, GNI dominates the whole analysis, accounting for 99.99% of variance via the first component, making the analysis useless. It is clearly very important to scale the variables when doing dimensionality reduction!

With standardized variables, we get two components accounting for ~70% of the total variance and possible interpretations.  

### Part 4: Interpretation of the components

Component 1 can be interpreted to represent (negative) prosperity. It is positively loaded by maternal mortality and adolescent birthrate, and negatively loaded by education and human development index. 

component 2 can be interpreted to represent gender equality, loaded by parliamentary representation of females and labor force participation ratio between the sexes. 

Components are orthogonal, indicating the they are, to a degree, independent factors withing countries. 


### Part 5: It's teatime!

```{r}

#read the data
tea <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv", stringsAsFactors = TRUE)

#check data, comment out
#view(tea)

#Let's use the perception variables from the last 12 columns of the data
tea_perc<-tea[,25:36]

#plot the responses
pivot_longer(tea_perc, cols = everything()) %>% 
  ggplot(aes(value)) + geom_bar()+ facet_wrap("name", scales = "free")+ theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))


#run MCA
mca <- MCA(tea_perc, graph = TRUE)

#plot the biplot
plot(mca, invisible=c("ind"), graph.type = "classic", habillage = "quali")

```





MCA was conducted on the variables detailing the perception of tea brands.

The two-component solution explains only quarter of the variance. The first dimension can be interpreted as positive/feminine vs negative/non-feminine definition of the product (DEFINITIONS). Second dimensions concerns health, excitability and vigour (EFFECTS).

More research is needed....





