# Chapter 6: Longitudinal data

load required libraries
```{r}
library(tidyverse)
library(psych)
library(corrplot)
library(ggplot2)
library(lme4)
library(Matrix)
library(car)
library(lmerTest)
```

### Part 1: Chapter 8 MABS analysis using RATS data

Read in the RATS data and convert to long form

```{r}
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header = TRUE, sep = '\t')

RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

RATSL <-  pivot_longer(RATS, cols = -c(ID, Group),
                       names_to = "weekdays", values_to = "rats") %>%
  arrange(weekdays) 

RATSL <-  RATSL %>% 
  mutate(time = as.integer(substr(weekdays,3,4)))

```


Plot the rat weight data by Time and Group

```{r}
ggplot(RATSL, aes(x = time, y = rats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:8, times=2)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$rats), max(RATSL$rats)))
```


Standardize by week and plot the standardized (by week) data by Time and Group

```{r}
RATSL <- RATSL %>%
  group_by(time) %>%
  mutate(stdrats = scale(rats)) %>%
  ungroup()


ggplot(RATSL, aes(x = time, y = stdrats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:8, times=2)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$stdrats), max(RATSL$stdrats)))
```


Groups 2 and 3 have much larger scores. There is an outlier in Group 2, a fat rat.


Plot mean and standard deviation by group. Used SD instead of SE since the group sizes differ.

```{r}
RATSS <- RATSL %>%
  group_by(Group, time) %>%
  summarise(mean =mean(rats), se = sd(rats)) %>%
  ungroup()

ggplot(RATSS, aes(x = time, y = mean, linetype = Group, colour = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.4)) +
  scale_y_continuous(name = "mean rat weight +/- sd(rat weight)")

```
Group 2 has large deviations, suggesting the presence of an outlier detected earlier.

Let's remove the outlier for exercise's sake. In real world, four observations per group is too small for any statistical operations, including checking for outliers.

Remove the outlier (ID=12) and make a boxplot per group.

```{r}
#Means with the outlier
RATSD2 <- RATSL %>%
 #filter(ID!=12)  %>%
 #filter(week > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(rats) ) %>%
  ungroup()

#Means without the ourlier
RATSD <- RATSL %>%
  filter(ID!=12)  %>%
 #filter(week > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(rats) ) %>%
  ungroup()

ggplot(RATSD, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean rat weight, days 1-64")
```

There are outliers in group 1 and group 3 as well but let's leave them be. Testing for group differences with the original outlier removed.

```{r}
t.test(mean ~ Group, data = filter(RATSD,Group!=3), var.equal = TRUE)
t.test(mean ~ Group, data = filter(RATSD,Group!=2), var.equal = TRUE)
t.test(mean ~ Group, data = filter(RATSD,Group!=1), var.equal = TRUE)
```

All differences are significant at p<0.01 level. Let's check without the ourlier removed:

```{r}
t.test(mean ~ Group, data = filter(RATSD2,Group!=3), var.equal = TRUE)
t.test(mean ~ Group, data = filter(RATSD2,Group!=2), var.equal = TRUE)
t.test(mean ~ Group, data = filter(RATSD2,Group!=1), var.equal = TRUE)
```

When the outlier in Group 2 remains, there difference between Group 2 and Group 3 is non-significant.

Create a new set of means without day 1, moving it to a separate baseline variable instead. With and without the outlier.

```{r}
rm(RATSD2)
RATSD2 <- RATSL %>%
  filter(time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(rats) ) %>%
  ungroup()

RATSB_WITHOUT <- RATSD2 %>%
  mutate(baseline = RATS$WD1)  %>%
   filter(ID!=12) 

RATSB_WITH <- RATSD2 %>%
  mutate(baseline = RATS$WD1) 
```

Fit the linear model with Group and baseline. Use car library to print ANOVA table. 

```{r}
fit <- lm(mean ~ Group + baseline, data = RATSB_WITHOUT)
Anova(fit, type="III")
```
When controlling for the first day baseline, there are still significant differences between groups at later days. Lets' check with the outlier included.


```{r}
fit <- lm(mean ~ Group + baseline, data = RATSB_WITH)
Anova(fit, type="III")
```

When the outlier in Group 2 is not removed, the difference between the groups becomes trending (p<0.10). Baseline still highly significant.


### Part 2: Chapter 9 MABS analysis using BPRS data

Read in the BPRS data and convert to long form
```{r}
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep =" ", header = T)

BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

BPRSL <-  pivot_longer(BPRS, cols = -c(treatment, subject),
                       names_to = "weeks", values_to = "bprs") %>%
  arrange(weeks) 

BPRSL <-  BPRSL %>% 
  mutate(week = as.integer(substr(weeks,5,5)))
```

Plot BPRS values per subjects, separate panels according to treatment (1 or 2)

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

Fit a regression model on the data with time and treatment as independent variables

```{r}
fit <- lm(bprs ~ treatment + week, data = BPRSL)
summary(fit)
```
Treatment does not seem to have an effect. Time (week) has a significant negative effect (p<0.001). The BPRS scores diminish over time.

Fit a linear mixed random intercept model with subject id (subject) as the random effect.

```{r}
bprs_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(bprs_ref)
```

Fit a linear mixed random slope and intercept model with subject id (subject) as the random effect with individual slopes for each individual. 


```{r}
bprs_ref2 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(bprs_ref2)
```
Test which model fits the data better

```{r}
anova(bprs_ref2, bprs_ref)
```
The random intercept and random slope model fits the data significantly better.

Fit a linear mixed random slope and intercept model with interaction between time and treatment.


```{r}
bprs_ref3 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(bprs_ref3)
```

Using lmertest package to get p values for the effects. It seems that the scores diminish as function of time (week) with a highly significant effect (p<0.001). Treatment does not affect scores, whereas time and treatment have trending interaction effect (p=0.08).


```{r}
anova(bprs_ref3, bprs_ref2)
```
Does not improve the fit significantly (p=0.08), but is an improvement still.


Plot the observed and fitted values with some jitter for readability.

```{r}
BPRSL$fitted <- fitted(bprs_ref3)  

ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_jitter(aes(color = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 2)) +
  scale_y_continuous(name = "Observed BPRS") +
  theme(legend.position = "top")

ggplot(BPRSL, aes(x = week, y = fitted, group = subject)) +
  geom_jitter(aes(color = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 2)) +
  scale_y_continuous(name = "Fitted BPRS") +
  theme(legend.position = "top")
```



It seems that the slope is slighty less negative for treatment group 2, as indicated by the trending interaction effect. Fit looks a bit aggressive, a lot of variation is unaccounted for. Main effect and it's direction are clearly observable.

The effect of time would perhaps better be modelled as quadratic, as some participants' scores, especially in group 2 ,seem to start rising at weeks 7 and 8. 
























